---
title: "STT 218 Data Analysis in R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{my-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
library(URStat218)
```

All notes in the vignettes of this package are adapted from Dr. Joseph Ciminelli's fall 2019 section of Statistics 218. For a more complete version of the material herein contained, please reference the notes, attend the lectures, or ask Dr. Ciminelli himself.  

## 2x2 Contingency Tables  
We begin by taking a contingency table on X and Y, where the rows are fixed:  

  
|X (Explanatory)|Y(Response)        |            |             |
|:--:           |:--:               |:--:        |:--:         |
|               |1                  |2           |  Total      |
|1              |   $n_{11}$        | $n_{12}$   |   $n_{1+}$  |
|2              |    $n_{21}$       |  $n_{22}$  |   $n_{2+}$  |

For this 2x2 table, we have two independent binomial proportions, one for each level of X:
$$n_{11} \sim Binom(n_{1+},p_{1|1})$$
$$n_{21} \sim Binom(n_{2+},p_{1|2})$$
Ultimately, we wish to test the significance of the difference in the two groups' proportions. If there is a significant difference, we conclude X and Y to be significantly associated, while the opposite leads to a correspondingly opposite conclusion.  
As we are testing for independence, with unknown probabilities $p_{1|1}$ and $p_{1|2}$, we test against a null hypothesis of $H_0: p_{1|1} =p_{1|2}$; that is, that the proportions are independent and equal. Of the methods available to us, we note that R has built-in support for a difference-of-proportions test. We discuss in greater detail Relative Risk and Odds Ratio tests, as this package creates functions for them.  
### Relative Risk:
